---
title: "Social Determinants of Health… and Wealth?"

author: C. Seth Lester, ASA (<a href="mailto:cslester@gmail.com">cslester@gmail.com</a>)<br>
date: 14 February 2023
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

_The following paper is a submission in reply to the [Society of Actuaries' Call for Papers](https://www.soa.org/research/opportunities/call-for-papers-list/) under the heading "[Aging and Retirement Issues for LGBTQ+ People – Second Invitation](https://www.soa.org/research/opportunities/call-for-essay-aging-retire-lgbtq/)"._

# Health and Wealth Are Heavily Socially Determined
For many years, people who identify as LGBTQ+ have endured discrimination from a wide variety of sources, such as from laws and regulations at all levels of government, or as a result of the ebb and flow of occupational attitudes over time, and even from members in our own community. 

For some LGBTQ+-identifying subpopulations, the level of discrimination we face in society is rapidly waning; for others, such as transgender individuals, there are still impactful and endemic stigmas in play which act as a barrier to these individuals from living full and rewarding lives. Barriers such as these have likely inhibited the ability of many people who identify as LGBTQ+ from enjoying the same protections of our social safety net – such as healthcare and retirement security systems – that are enjoyed by non-LGBTQ+-identifying people.

It is difficult today for analysts to credibly estimate how retirement-related outcomes differ between LGBTQ+ and non-LGBTQ+ populations, or how to quantify the disparities of outcomes among intersectional subpopulations that exist today within LGBTQ+ communities. One primary reason motivating this difficulty stems from the lack of available data that can help guide our policymaking apparatus towards better regulations and protections for members of our society who are more likely to be marginalized, stigmatized, or disadvantaged by systemic factors. This includes LGBTQ+ people, of course, but heavily intersects with people of color, women, religious minorities, immigrants, disabled individuals, and countless other populations of interest that comprise the greater whole of the nation.

In the healthcare space, there is ample energy surrounding the important work being done by organizations, both private and public, to better understand the impacts of what are known as Social Determinants of Health – “the conditions in the environments where people are born, live, learn, work, play, worship, and age that affect a wide range of health, functioning, and quality-of-life outcomes and risks” (definition lifted from the [US Dept. of Health and Human Services](https://health.gov/healthypeople/priority-areas/social-determinants-health).

Social Determinants of Health (SDoH) can provide strong causal evidence for why certain subpopulations of our society are marginalized or disadvantaged by systemic factors. Consequently, population health professionals and actuaries alike have an interest in better understanding the relationships between SDoH and access to quality healthcare that is both affordable and can be sustainably provided by the care delivery system.

It is therefore reasonable to suspect that disparities in retirement-related outcomes between LGBTQ+ and non-LGBTQ+ populations are also highly dependent upon many of the same Social Determinants of Health (SDoH) factors that heavily influence healthcare outcomes, experiences, and costs.

In order to investigate this idea, I set out to explore the publicly-available [Medical Expenditure Panel Survey](https://meps.ahrq.gov/mepsweb/) (MEPS) datasets to determine if there were useful data available for distilling connections between retirement-related outcomes, many of which depend extensively on healthcare outcomes, and the lived socioeconomic experiences often encountered by LGBTQ+ individuals.

# About MEPS
The Medical Expenditure Panel Survey has been administered since 1996, and according the [Agency for Healthcare Research and Quality](https://www.ahrq.gov/) (AHRQ), which is the government agency at MEPS’ helm, MEPS is a collection of “data on the specific health services that Americans use, how frequently they use them, the cost of these services, and how they are paid for, as well as data on the cost, scope, and breadth of health insurance held by and available to U.S. workers” ([Survey Background, MEPS Homepage](https://meps.ahrq.gov/mepsweb/about_meps/survey_back.jsp)).

On the [MEPS Github page](https://github.com/HHS-AHRQ/MEPS), you’ll find several excellent code examples written in both SAS and R which I used to familiarize myself with the layout of the data. Furthermore, the MEPS staff regularly conducts well-crafted online training seminars.

As MEPS is a survey and therefore the data is either self-reported or imputed, there are some situations in which you’ll find the data isn’t perfect. And furthermore, MEPS provides [some basic ground rules](https://meps.ahrq.gov/survey_comp/precision_guidelines.shtml) about making inferences with the data, which include restrictions and guidance concerning how much you are permitted to slice and dice the data before your unweighted sample of respondents is too small. I discuss this guidance further in a subsequent section.

# My Approach: MEPS and LGBTQ+ Individuals
The MEPS questionnaire is broad and comprehensive, and includes questions ranging from basic demography facts to employment status, food stamp usage, diagnoses of key chronic conditions, who pays for the respondent’s insurance, what medications they filled, medical procedures they received, even down to the ethnicity of their providers.

However, one question that is glaringly absent from the panel survey questionnaire is any mention of the respondent’s LGBTQ+ identity. There are very good reasons to suspect adding these questions would impose unnecessary sample biases. 

At any rate, because MEPS does provide a way to identify respondents in the survey who are spouses of one another, and because MEPS provides the self-reported sex of each respondent, one can use this approach to create a dataset of all married respondents, and then easily label which are in same-sex marriages and which are not in same-sex marriages.

The R code that performs this analysis and prepares the data visualizations that follow is available on my personal Github site (the repository and content which you are currently reading).

# Caveats and Housekeeping
In the following infographics, I have used the 2019 datasets available on the MEPS website. Note that these are actual populations estimates based on the weighted survey design, so the population counts and dollar amounts below are intended to be representative of the US population, in context.

Next, please take care to note that MEPS recommends [not displaying estimates of any kind with a RSE (relative standard error) of .5 or higher](https://meps.ahrq.gov/survey_comp/precision_guidelines.shtml) due to the high degree of sampling error. Therefore, in the following images, take care to note that error bars extending more than 50% into the shaded bar should not be taken at face value. Conversely, error bars extending no more than 30% into the shaded bar can be regarded as sufficiently precise.

Finally, I want to reiterate that while the distinction between sex and gender is not lost on me (in particular in the context of analyzing causal connections between SDoH and healthcare outcomes), I must point out that the variable called “SEX” in the data is indeed, according to the MEPS handbook, intended to express the respondent’s gender. The handbook also points out that sometimes this value is assigned based on the name of the respondent or based on the relationship to other respondents in the same family or dwelling unit, which is obviously not ideal.

For consistency’s sake, I continue to refer to the variable in the context of sex and not gender in the visualizations below. This also means it is possible that some of these “same-sex” and “non-same-sex” couples could potentially include transgender individuals, but I will proceed with the materiality assumption that this does not occur in the data very frequently.

# Exploring the Demographic Composition of Same-Sex Marriages in the US
In the following infographics, I attempt to better understand the demographic composition of individuals in same-sex marriages by exploring the relationships between age, sex, income level, race, and insurance coverage status between the same-sex and non-same-sex marriage cohorts.

```{r preprocessing, include=FALSE}

# Invoke required libraries
library(survey)
library(foreign)
library(haven)
library(viridis)
library(tidyverse)
library(MEPS)

# Set options to deal with lonely psu
options(survey.lonely.psu='adjust')

# Read in data from FYC file --------------------------------------------------

# fyc20 = read_MEPS(year = 2020, type = "FYC") # 2020 FYC
fyc19 = read_MEPS(year = 2019, type = "FYC") # 2019 FYC
# fyc18 = read_MEPS(year = 2018, type = "FYC") # 2018 FYC

# Identify SSCs - follow instructions in section 3.5 to create one family per record, but 
# consolidate tibbles within each record to have individual-level details.

count_family_size <- function(df) {
  # First get list of PIDs
  pid_list <- df %>% 
    distinct(PID) %>% 
    nrow()
}

identify_same_sex_spouses <- function(df) {
  # Join PIDs table onto itself by SPOUIDs to determine spousal match-ups at the PID level
  # Then you can identify if the "left hand" and "right hand" sexes match.
  
  pid_list <- df %>% 
    distinct(PID, SEX, SPOUIDYY)
  
  pid_join <- pid_list %>% 
    inner_join(pid_list, by=c("PID"="SPOUIDYY")) %>% 
    mutate(sex_of_spouse = SEX.y,
           same_sex_lgl = SEX.x == SEX.y) %>% 
    select(PID, sex_of_spouse, same_sex_lgl)
  
  pid_join
}

count_married_individuals <- function(df) {
  pid_list <- df %>% 
    filter(MARRYYYX == 1) %>% 
    distinct(PID) %>% 
    nrow()
}

count_married_individuals_in_ssc <- function(df) {
  count <- df %>% 
    filter(same_sex_lgl == T) %>% 
    nrow()
}

# fyc20x  <- fyc20 %>% 
#   select(DUID, PID, DUPERSID, SEX, AGE20X, VARSTR, VARPSU, PERWT20F, MARRY20X, SPOUID20, FAMIDYR, FAMWT20F, FAMRFPYR) %>% 
#   filter(FAMWT20F > 0) %>%      # Per instruction in section 3.5
#   mutate(DUIDFAMY = stringr::str_c(DUID, FAMIDYR)) %>% # Per instruction in section 3.5
#   rename(PERWTYYF=PERWT20F,
#          MARRYYYX=MARRY20X,
#          SPOUIDYY=SPOUID20,
#          FAMWTYYF=FAMWT20F,
#          AGEYYX=AGE20X) %>% 
#   group_by(DUIDFAMY) %>%  
#   tidyr::nest() %>% 
#   ungroup() %>% 
#   mutate(family_size = purrr::map_int(data, count_family_size)) %>% 
#   mutate(married_individuals = purrr::map_int(data, count_married_individuals)) %>% 
#   mutate(spouse_matrix = purrr::map(data, identify_same_sex_spouses)) %>% 
#   mutate(married_individuals_in_ssc = purrr::map_int(spouse_matrix, count_married_individuals_in_ssc))

fyc19x  <- fyc19 %>% 
  select(DUID, PID, DUPERSID, SEX, AGELAST, INSCOV19, REGION19, RACETHX, POVCAT19, TOTEXP19, VARSTR, VARPSU, PERWT19F, MARRY19X, SPOUID19, FAMIDYR, FAMWT19F, FAMRFPYR) %>% 
  filter(FAMWT19F > 0) %>%      # Per instruction in section 3.5
  mutate(DUIDFAMY = stringr::str_c(DUID, FAMIDYR)) %>% # Per instruction in section 3.5
  rename(PERWTYYF=PERWT19F,
         MARRYYYX=MARRY19X,
         SPOUIDYY=SPOUID19,
         FAMWTYYF=FAMWT19F,
         AGEYYX=AGELAST,
         REGIONYY=REGION19,
         INSCOVYY=INSCOV19,
         TOTEXPYY=TOTEXP19,
         POVCATYY=POVCAT19) %>% 
  group_by(DUIDFAMY) %>%  
  tidyr::nest() %>% 
  mutate(family_size = purrr::map_int(data, count_family_size)) %>% 
  mutate(married_individuals = purrr::map_int(data, count_married_individuals)) %>% 
  mutate(spouse_matrix = purrr::map(data, identify_same_sex_spouses)) %>% 
  mutate(married_individuals_in_ssc = purrr::map_int(spouse_matrix, count_married_individuals_in_ssc))

fyc19x_flat <- fyc19x %>% 
  mutate(final_data = purrr::map2(data, spouse_matrix, ~ .x %>% left_join(.y, by=c("PID"="PID")))) %>% 
  select(-data, -spouse_matrix) %>% 
  group_by(DUIDFAMY, family_size, married_individuals, married_individuals_in_ssc) %>% 
  tidyr::unnest(final_data) %>% 
  ungroup() %>% 
  tidyr::replace_na(list(same_sex_lgl=F)) %>% 
  filter(AGEYYX > 0) %>% 
  filter(MARRYYYX == 1 & !is.na(sex_of_spouse)) %>% 
  mutate(unwt_persons = 1) %>% 
  mutate(AGE_GRP_2 = if_else(AGEYYX >= 65, "65+", "Under 65")) %>% 
  mutate(AGE_GRP_5 = case_when(AGEYYX < 5 ~ "Under 5",
                               AGEYYX >= 5 & AGEYYX <= 17 ~ "5 - 17",
                               AGEYYX >= 18 & AGEYYX <= 44 ~ "18 - 44",
                               AGEYYX >= 45 & AGEYYX <= 64 ~ "45 - 64",
                               AGEYYX >= 65 ~ "65+",
                               T ~ as.character(AGEYYX))) %>%
  mutate(INSCOV_DSC = case_when(INSCOVYY == 1 ~ "Any Private",
                                INSCOVYY == 2 ~ "Public Only",
                                INSCOVYY == 3 ~ "Uninsured",
                                T ~ as.character(INSCOVYY))) %>% 
  mutate(UNINS_FLG = case_when(INSCOVYY == 3 ~ "Uninsured",
                               INSCOVYY != 3 ~ "Insured",
                               T ~ as.character(INSCOVYY))) %>% 
  mutate(REG_DSC = case_when(REGIONYY == -1 ~ "Inapplicable",
                             REGIONYY == 1 ~ "Northeast",
                             REGIONYY == 2 ~ "Midwest",
                             REGIONYY == 3 ~ "South",
                             REGIONYY == 4 ~ "West",
                             T ~ as.character(REGIONYY))) %>% 
  mutate(RACE_DSC = case_when(RACETHX == 1 ~ "Hispanic",
                              RACETHX == 2 ~ "Non-Hispanic White Only",
                              RACETHX == 3 ~ "Non-Hispanic Black Only",
                              RACETHX == 4 ~ "Non-Hispanic Asian Only",
                              RACETHX == 5 ~ "Non-Hispanic Other Race or Multiple Race",
                              T ~ as.character(RACETHX))) %>% 
  mutate(POVCAT_DSC = case_when(POVCATYY == 1 ~ "Poor / Negative",
                                POVCATYY == 2 ~ "Near Poor",
                                POVCATYY == 3 ~ "Low Income",
                                POVCATYY == 4 ~ "Middle Income",
                                POVCATYY == 5 ~ "High Income",
                                T ~ as.character(POVCATYY))) %>% 
  mutate(HAS_EXP = if_else(TOTEXPYY > 0, "Had Healthcare Expenses", "Did Not Have Healthcare Expenses")) %>% 
  mutate(SEX_DSC = if_else(SEX == 1, "Male", if_else(SEX == 2, "Female", "Unknown"))) %>% 
  mutate(SSC_DSC = if_else(same_sex_lgl == T, "Individuals in Same-Sex Marriages", "Individuals Not in Same-Sex Marriages"))

# Define the survey design

mepsdsgn = svydesign(
  id = ~VARPSU,
  strata = ~VARSTR,
  weights = ~PERWTYYF,
  data = fyc19x_flat,
  nest = TRUE)


```


```{r pressure, echo=FALSE}

fyc19x_age_sex <- svyby(~unwt_persons, by = ~AGE_GRP_5 + SEX_DSC + SSC_DSC, FUN = svytotal, design = mepsdsgn) 

# visualize these demos to make sure we're comparing apples-to-apples

fyc19x_demo_gg1 <- fyc19x_age_sex %>% 
  ggplot(mapping = aes(x = AGE_GRP_5,
                       fill = SEX_DSC,
                       y = unwt_persons)) +
  scale_fill_viridis(discrete=T, begin=0.5, end = 1) +
  geom_bar(stat="identity", position="dodge") +
  geom_errorbar(aes(ymin=unwt_persons-se, ymax=unwt_persons+se), width=.2,
                position=position_dodge(.9)) +
  theme_bw() +
  scale_y_continuous(labels = scales::comma) +
  facet_wrap(~ SSC_DSC, ncol = 2, scales="free_y", labeller = label_wrap_gen(width = 18, multi_line = TRUE)) +
  labs(title="Estimated Number of Married Individuals",
       subtitle="By Marriage Classification and Sex",
       x = "Age Grouping",
       y = "Estimate of Individuals (Survey Weighted)",
       fill = "Sex",
       caption = "MEPS 2019 Full Year Consolidated Data File\n(https://meps.ahrq.gov/)") +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 3))

fyc19x_demo_gg1

fyc19x_age_cov <- svyby(~unwt_persons, by = ~AGE_GRP_5 + INSCOV_DSC + SSC_DSC, FUN = svytotal, design = mepsdsgn) 

fyc19x_demo_gg2 <- fyc19x_age_cov %>% 
  ggplot(mapping = aes(x = AGE_GRP_5,
                       fill = INSCOV_DSC,
                       y = unwt_persons)) +
  scale_fill_viridis(discrete=T, begin=0.5, end = 1) +
  geom_bar(stat="identity", position="dodge") +
  geom_errorbar(aes(ymin=unwt_persons-se, ymax=unwt_persons+se), width=.2,
                position=position_dodge(.9)) +
  theme_bw() +
  scale_y_continuous(labels = scales::comma) +
  facet_wrap(~ SSC_DSC, ncol = 2, scales="free_y", labeller = label_wrap_gen(width = 18, multi_line = TRUE)) +
  labs(title="Estimated Number of Married Individuals",
       subtitle=" By Marriage Classification and Health Plan Coverage",
       x = "Age Grouping",
       y = "Estimate of Individuals (Survey Weighted)",
       fill = "Health Plan Coverage",
       caption = "MEPS 2019 Full Year Consolidated Data File\n(https://meps.ahrq.gov/)") +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 3))

fyc19x_demo_gg2

fyc19x_age_race <- svyby(~unwt_persons, by = ~AGE_GRP_5 + RACE_DSC + SSC_DSC, FUN = svytotal, design = mepsdsgn) 

fyc19x_demo_gg3 <- fyc19x_age_race %>% 
  ggplot(mapping = aes(x = AGE_GRP_5,
                       fill = RACE_DSC,
                       y = unwt_persons)) +
  scale_fill_viridis(discrete=T, begin=0.5, end = 1) +
  geom_bar(stat="identity", position="dodge") +
  geom_errorbar(aes(ymin=unwt_persons-se, ymax=unwt_persons+se), width=.2,
                position=position_dodge(.9)) +
  theme_bw() +
  scale_y_continuous(labels = scales::comma) +
  facet_wrap(~ SSC_DSC, ncol = 2, scales="free_y", labeller = label_wrap_gen(width = 18, multi_line = TRUE)) +
  labs(title="Estimated Number of Married Individuals",
       subtitle="By Marriage Classification and Race, Ethnicity",
       x = "Age Grouping",
       y = "Estimate of Individuals (Survey Weighted)",
       fill = "Race and Ethnicity",
       caption = "MEPS 2019 Full Year Consolidated Data File\n(https://meps.ahrq.gov/)") +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 3))

fyc19x_demo_gg3

fyc19x_age_povcat <- svyby(~unwt_persons, by = ~AGE_GRP_5 + POVCAT_DSC + SSC_DSC, FUN = svytotal, design = mepsdsgn) 

fyc19x_demo_gg4 <- fyc19x_age_povcat %>% 
  mutate(povcat_ordered = fct_relevel(POVCAT_DSC, "High Income", "Middle Income", "Low Income", "Near Poor", "Poor / Negative")) %>% 
  ggplot(mapping = aes(x = AGE_GRP_5,
                       fill = povcat_ordered,
                       y = unwt_persons)) +
  scale_fill_viridis(discrete=T, begin=0.5, end = 1) +
  geom_bar(stat="identity", position="dodge") +
  geom_errorbar(aes(ymin=unwt_persons-se, ymax=unwt_persons+se), width=.2,
                position=position_dodge(.9)) +
  theme_bw() +
  scale_y_continuous(labels = scales::comma) +
  facet_wrap(~ SSC_DSC, ncol = 2, scales="free_y", labeller = label_wrap_gen(width = 18, multi_line = TRUE)) +
  labs(title="Estimated Number of Married Individuals",
       subtitle="By Marriage Classification and Income",
       x = "Age Grouping",
       y = "Estimate of Individuals (Survey Weighted)",
       fill = "Income Level",
       caption = "MEPS 2019 Full Year Consolidated Data File\n(https://meps.ahrq.gov/)") +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 3))

fyc19x_demo_gg4

```

One surprising piece of information I was able to learn from these images is precisely how stark the difference is between the percentage of non-Hispanic white individuals in same-sex marriages vs. non-same-sex marriages, which potentially could indicate a greater deal of social acceptance of same-sex marriages among non-Hispanic white populations. 

This is an important point of discussion as it illustrates the need for researchers and policymakers to better understand the idea of intersectionality within the many subpopulations within LGBTQ+ groups. As marriage tends to confer financial security and stabilization benefits, to both participants in the couple, then it follows that LGBTQ+ individuals who are white are more likely to access these benefits than their non-white LGBTQ+ counterparts, which can contribute to additional disparities along racial lines which are fully contained within the LGBTQ+ population.

# Visualizing Total Healthcare Expenditures
It is also the case that LGBTQ+ individuals can possess unique healthcare needs or face different barriers to access to healthcare, both of which can move the needle on the cost of healthcare – particularly in retirement, in which individuals are more likely to face one or more costly chronic conditions such as heart disease, diabetes, or asthma. 

To better determine if there are relationships between these demographic factors and the cost of care – particularly for older age groups, I prepared the following series of visualizations below that examine the average individual expenditures broken out on the same demographic features as the visualizations above.


```{r echo=FALSE}

fyc19x_age_sex_exp <- svyby(~TOTEXPYY, by = ~AGE_GRP_5 + SEX_DSC + SSC_DSC, FUN = svymean, design = mepsdsgn) 

fyc19x_demo_gg1e <- fyc19x_age_sex_exp %>% 
  ggplot(mapping = aes(x = AGE_GRP_5,
                       fill = SEX_DSC,
                       y = TOTEXPYY)) +
  scale_fill_viridis(discrete=T, begin=0.5, end = 1) +
  geom_bar(stat="identity", position="dodge") +
  geom_errorbar(aes(ymin=TOTEXPYY-se, ymax=TOTEXPYY+se), width=.2,
                position=position_dodge(.9)) +
  theme_bw() +
  scale_y_continuous(labels = scales::dollar) +
  facet_wrap(~ SSC_DSC, ncol = 2, labeller = label_wrap_gen(width = 18, multi_line = TRUE)) +
  labs(title="Estimated Healthcare Expenditures",
       subtitle="By Marriage Classification and Sex",
       x = "Age Grouping",
       y = "Individual Healthcare Spend (Survey Weighted)",
       fill = "Sex",
       caption = "MEPS 2019 Full Year Consolidated Data File\n(https://meps.ahrq.gov/)") +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 3))

fyc19x_demo_gg1e

fyc19x_age_cov_exp <- svyby(~TOTEXPYY, by = ~AGE_GRP_5 + INSCOV_DSC + SSC_DSC, FUN = svymean, design = mepsdsgn) 

fyc19x_demo_gg2e <- fyc19x_age_cov_exp %>% 
  ggplot(mapping = aes(x = AGE_GRP_5,
                       fill = INSCOV_DSC,
                       y = TOTEXPYY)) +
  scale_fill_viridis(discrete=T, begin=0.5, end = 1) +
  geom_bar(stat="identity", position="dodge") +
  geom_errorbar(aes(ymin=TOTEXPYY-se, ymax=TOTEXPYY+se), width=.2,
                position=position_dodge(.9)) +
  theme_bw() +
  scale_y_continuous(labels = scales::comma) +
  facet_wrap(~ SSC_DSC, ncol = 2, labeller = label_wrap_gen(width = 18, multi_line = TRUE)) +
  labs(title="Estimated Healthcare Expenditures",
       subtitle="By Marriage Classification and Health Plan Coverage",
       x = "Age Grouping",
       y = "Individual Healthcare Spend (Survey Weighted)",
       fill = "Health Plan Coverage",
       caption = "MEPS 2019 Full Year Consolidated Data File\n(https://meps.ahrq.gov/)") +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 3))

fyc19x_demo_gg2e

fyc19x_age_race_exp <- svyby(~TOTEXPYY, by = ~AGE_GRP_5 + RACE_DSC + SSC_DSC, FUN = svymean, design = mepsdsgn) 

fyc19x_demo_gg3e <- fyc19x_age_race_exp %>% 
  ggplot(mapping = aes(x = AGE_GRP_5,
                       fill = RACE_DSC,
                       y = TOTEXPYY)) +
  scale_fill_viridis(discrete=T, begin=0.5, end = 1) +
  geom_bar(stat="identity", position="dodge") +
  geom_errorbar(aes(ymin=TOTEXPYY-se, ymax=TOTEXPYY+se), width=.2,
                position=position_dodge(.9)) +
  theme_bw() +
  scale_y_continuous(labels = scales::dollar) +
  facet_wrap(~ SSC_DSC, ncol = 2, labeller = label_wrap_gen(width = 18, multi_line = TRUE)) +
  labs(title="Estimated Healthcare Expenditures",
       subtitle="By Marriage Classification and Race, Ethnicity",
       x = "Age Grouping",
       y = "Individual Healthcare Spend (Survey Weighted)",
       fill = "Race and Ethnicity",
       caption = "MEPS 2019 Full Year Consolidated Data File\n(https://meps.ahrq.gov/)") +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 3))

fyc19x_demo_gg3e

fyc19x_age_povcat_exp <- svyby(~TOTEXPYY, by = ~AGE_GRP_5 + POVCAT_DSC + SSC_DSC, FUN = svymean, design = mepsdsgn) 

fyc19x_demo_gg4e <- fyc19x_age_povcat_exp %>% 
  mutate(povcat_ordered = fct_relevel(POVCAT_DSC, "High Income", "Middle Income", "Low Income", "Near Poor", "Poor / Negative")) %>% 
  ggplot(mapping = aes(x = AGE_GRP_5,
                       fill = povcat_ordered,
                       y = TOTEXPYY)) +
  scale_fill_viridis(discrete=T, begin=0.5, end = 1) +
  geom_bar(stat="identity", position="dodge") +
  geom_errorbar(aes(ymin=TOTEXPYY-se, ymax=TOTEXPYY+se), width=.2,
                position=position_dodge(.9)) +
  theme_bw() +
  scale_y_continuous(labels = scales::dollar) +
  facet_wrap(~ SSC_DSC, ncol = 2, labeller = label_wrap_gen(width = 18, multi_line = TRUE)) +
  labs(title="Estimated Healthcare Expenditures",
       subtitle="By Marriage Classification and Income",
       x = "Age Grouping",
       y = "Individual Healthcare Spend (Survey Weighted)",
       fill = "Income Level",
       caption = "MEPS 2019 Full Year Consolidated Data File\n(https://meps.ahrq.gov/)") +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 3))

fyc19x_demo_gg4e

```

# Conclusion
Due to the lack of data currently available, we must continue to be creative about how we source information that helps build policy and products to better strengthen the lives of LGBTQ+ individuals who are currently enjoying retirement, as well as those who will be retiring soon.

While we’ve only just skimmed the surface of what MEPS has to offer, MEPS provides an excellent cross section of data collected in a well-designed survey framework that can aid key decision makers as they go about analyzing the impacts that Social Determinants of Health have on retirement-related outcomes. 

Some examples of other potentially interesting variables in the context of examining retirement-related outcomes alongside the ones I’ve shown in this paper include:

```
MILDIF31    DIFFICULTY WALKING A MILE - RD 3/1 
MIAGED      AGE OF DIAGNOSIS-HEART ATTACK(MI) 
FAMSZEYR 		SIZE OF RESPONDING ANNUALIZED FAMILY 
WHTLGSPK 		WHAT LANGUAGE SPOKEN OTHER THAN ENGLISH 
DDNWRK19 		# DAYS MISSED WORK DUE TO ILL/INJ 2019 
ADRESP42 		SAQ 12 MOS: DR SHOWED RESPECT 
PROBPY42 		FAMILY HAVING PROB PAYING MEDICAL BILLS 
```

And while the particular method I’ve demonstrated in this paper for identifying LGBTQ+ individuals and experiences certainly does have its flaws, we have to do the best we can with the information we have – and publicly available data on the needs of LGBTQ+ individuals at or near retirement age is frustratingly scarce.

Finally, this sort of work encourages discussions that broaden our understanding of how systemic disparities act with intersectionality among other groups and subpopulations. And while understanding the concept of intersectionality is one thing, being able to quantify the result of intersectional disparate outcomes, even if the data isn’t perfect, is entirely more useful.


